An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-05-04 cal_demo/ms.tex
cleaning up derivation of cal, retaining noise throughout

Because the gains must be estimated in real time, we will need to use a finite stream of data to 
form and estimate which will be applied to the subsequent stream of data. Therefore the 
process of finding a solution is an iterative one. We will use parenthetical superscripts to 
denote the calibration loop number. For example, we will assume we have already formed an 
estimate from $n$ loops, $g^{(n)}_a$. We will use these estimates to calibrate the next stream 
of data in order to form an updated estimate, $g^{(n+1)}_a$.

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Cna}
\Cna \equiv \left<\Er{a}(t) E'^*(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, we find,
\begin{align}\label{eq:cna}
\Cna & = \left<g^T_a \Er{a}^T(t) \frac{1}{\Nant} \sum_b \frac{g^{T*}_b}{g^{*(n)}_b}\Er{b}^{T*}(t) e^{-2\pi i \spix \cdot \rb} \beamtheta^*_b(\spix)\right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b \frac{g^{T*}_b}{g^{*(n)}_b} \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left<\Er{a}^T \Er{b}^{*T} \right>_t \nonumber \\
& = \frac{g^T_a}{\Nant} \sum_b \frac{g^{T*}_b}{g^{*(n)}_b} \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \V^T_{ab}
\end{align}
where in the second step we group time-dependent terms, and in the third we define the true 
visibilities as the correlation between true antenna electric field measurements. It is easy to see 
from here that the net effect of including the receiver noise term, $\widetilde{n}_a(f,t)$, will 
result in added noise on the true visibilities, including a bias on the auto-correlation terms. In 
principle a bias can result on any visibility from any noise correlations between antennas, but 
the implementation included in the EPIC software package restricts the noise term to include 
only the auto-correlation noise bias, and assumes baseline dependent noise has zero mean.

>>>>

Because the gains must be estimated in real time, we will need to use a finite stream of data to 
form an estimate which will be applied to the subsequent stream of data. Therefore the 
process of finding a solution is an iterative one. We will use parenthetical superscripts to 
denote the calibration loop number. For example, we will assume we have already formed an 
estimate from $n$ loops, $g^{(n)}_a$. We will use these estimates to calibrate the next stream 
of data in order to form an updated estimate, $g^{(n+1)}_a$.

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the complex conjugate of an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Cna}
\Cna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
the superscript $^*$ indicates complex conjugation,
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, 
and substituting the $g^{(n)}_b$ for $g'_b$ to indicate the specific loop number,
we find,
\begin{align}\label{eq:cna}
\Cna & = \left<\left(g_a E_a^I(t)+n_a(t)\right) \frac{1}{\Nant} \right. \nonumber \\
& \qquad \left. \times \sum_b \frac{1}{g^{*(n)}_b}\left(g^*_bE_b^{I*}(t)+n^*_b(t)\right) e^{-2\pi i \spix \cdot \rb} \mathcal{W}^*_b(\spix)\right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left<\left(g_aE_a^I(t)+n_a(t)\right)\left(g^*_bE_b^{*I}+n^*_b(t)\right) \right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left(g_a g^*_b \left<E^I_a(t)E^{*I}_b(t)\right>_t + \left<n_a(t) n^*_b(t)\right>_t\right) \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left(g_ag^*_bV^I_{ab}+n_{ab}\right)
\end{align}
where in the second step we group time-dependent terms, 
in the third step we assume the noise does not correlate with the ideal electric field 
measurements, and in the final step we introduce the baseline noise $n_{ab}\equiv
\left<n_a(t)n^*_b(t)\right>_t$, and ideal visibilities as the correlation between ideal
electric field measurements, $V^I_{ab} \equiv\left<E^I_a(t)E^{*I}_b(t)\right>_t$.
To be clear, visibilities are never directly formed from the data. Rather it is convenient to
express $\Cna$ in terms of the ideal visibilities, which we can model when solving for the
gains below.

\textcolor{red}{need a brief comment on the noise term}
